Parsing Postfix log files

### 2009-03-14 doc/logparser-implementation.tex
Dealt with some XXX items.

    \item [CONNECT] Handle a remote client connecting: create a new
        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
        data captured by the rule (\acronym{IP} address and hostname of the
        client).  If a connection already exists it is treated as a symptom
        of a bug, and the action issues a warning containing the full
        contents of the existing connection plus the log line that has
        just been parsed.

@@

XXX DESCRIBE HOW TO GO ABOUT ADDING A NEW RULE\@.

@@

XXX SHOULD I HAVE A LIST OF HOW MANY COMPLICATIONS ARE HANDLED BY EACH
ACTION\@?

@@

@@

XXX WOULD IT BE BETTER TO BEGIN THE EXPLANATIONS ON THE FOLLOWING LINE\@?

@@

    \item XXX IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LOG LINES\@?
        There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
        log lines for the mail, with no additional log lines.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop\glsadd{mail
loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
Postfix has had a chance to try to deliver it.  In this case the recipient
address will not have been logged, so a dummy recipient address needs to be
added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  This
complication has been observed in XXX/\numberOFlogFILES{} log files, but
typically when it does occur there will be many instances of it, closely
grouped.

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their source is logged.  To
deal with this the COMMIT action maintains a cache of recently committed
bounce notification mails, which the BOUNCE action subsequently checks if
the bounce mail is not already in the state tables.  If the queueid exists
in the cache, and its start time is less than ten seconds before the
timestamp of the bounce log line, it is assumed that the bounce
notification mail has already been processed and the BOUNCE action does not
create one.  If the queueid exists it is removed from the cache, because it
has either just been used or it is too old to use in future.  Whether the
BOUNCE action creates a mail or finds an existing mail in the state tables,
it flags the mail as having been seen by the BOUNCE action; if this flag is
present the COMMIT action will not add the mail to the cache of recent
bounce notification mails.\footnote{This is not required to correctly deal
with the complication, but is an optimisation to reduce the parser's memory
usage; on the occasions the author has observed this complication occurring
there have been a huge number of bounce notification mails generated --- if
every bounce notification mail was cached it would dramatically increase
the memory requirements of the parser.}  The cache of bounce notification
mails will be pruned whenever the parser's state is saved, though if the
size of the cache ever becomes a problem it could be pruned periodically to
keep the size in check.  XXX EXPLAIN PRUNING\@.

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \acronym{DNSBL}, a new policy server, or new
        administrator-defined rejection messages require new rules.

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not be a queueid
        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
        evaluated for the first recipient, so the log line cannot be
        associated with the correct mail.  There is guaranteed to be a
        queueid when the DATA command has been reached, and thus the
        queueid will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

        Logging the HELO hostname in this fashion also partially prevents
        the complication described in \sectionref{Mail deleted before
        delivery is attempted} from occurring, but only when there is a
        single recipient; in that case the recipient address will be logged
        also, but when there are multiple recipients no addresses are
        logged.  It is also possible to warn for every recipient,
        preventing the complication in \sectionref{Mail deleted before
        delivery is attempted} entirely.  XXX ADD A PARAGRAPH EXPLAINING
        HOW\@.

@@

    \item The parser does not distinguish between log files produced by
        different sources when parsing; all results will be saved to the
        same database.  This may be viewed as an advantage, as log files
        from different sources can be combined in the same database, or it
        may be viewed as a limitation as there is no facility to
        distinguish between log files from different sources in the same
        database.  If the results of parsing log files from different
        sources must remain separate, the parser can easily be instructed
        to use a different database to store the results in.

>>>>

    \item [CONNECT] Handle a remote client connecting: create a new
        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
        data captured by the rule's regex (\acronym{IP} address and
        hostname of the client).  If a connection already exists it is
        treated as a symptom of a bug in \parsername{}, and the action will
        issue a warning containing the full contents of the existing
        connection plus the log line that has just been parsed.

@@

To add a new rule a new row must be added to the rules table in the
database, containing all the required attributes: action, name,
description, program, and regex.  The name and description fields should be
set based on the meaning of the log line, to help others understand which
log lines this rule will recognise; the program will be obvious from the
unrecognised log lines.  The action depends on the what the log line
represents, e.g.\ a delivery rejection, a mail being delivered, some useful
information, or something else; examine the list of actions in
\sectionref{actions in detail in implementation} to determine the correct
one.  The regex will be constructed based on the log line, but see below
for a tool to ease the process.

There are other attributes that may be required in a rule:
connection\_data, result\_data, priority, or restriction\_name.  In general
it will only become clear that connection\_data or result\_data are
required when \parsername{} warns about an entry in the connections or
results tables that is missing some required fields, because values for the
missing fields are not present in any of the log lines.  E.g.\ the rule
recognising the \daemon{pickup} component processing a mail sets the
client\_hostname to \texttt{localhost} and the client\_ip to
\texttt{127.0.0.1}, because the mail originates on the local server.  If
the new rule deliberately uses the architecture's overlapping rules feature
it will be obvious that the priority field needs to be set, on this rule and
possibly others; the priority field may be needed on unintentionally
overlapping rules too, but that is more difficult to determine.  Finally
the restriction\_name field should be set when the rule's action is
\texttt{DELIVERY\_REJECTED}; hopefully the restriction in question will be
clear from the content of the log line.

@@

@@

\newpage{} % XXX ADDED BECAUSE OF FLIP-FLOPPING LAYOUT.

@@

% This will put the text on the line following the item name, if the
% enumitem package is loaded.
%\begin{description}[style=nextline]
\begin{description}

@@

    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
        log lines for the mail, with no additional log lines.

@@

Postfix logs the recipient address when delivery of a mail is attempted, so
if delivery has yet to be attempted the parser cannot determine the
recipient address or addresses.  This is a problem when mail is arriving
faster than Postfix can attempt delivery, and the administrator deletes
some of the mail (because it is the result of a mail loop\glsadd{mail
loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
Postfix has had a chance to try to deliver it.  In this case the recipient
address will not have been logged, so a dummy recipient address needs to be
added, as every mail is required by the database schema
(\sectionref{results table}) to have at least one recipient.  This
complication has been observed in XXX/\numberOFlogFILES{} log files, but
typically when it does occur there were many instances of it, closely
grouped.

This lack of information cannot easily be overcome: it is trivial to log a
warning for every recipient accepted, but when the warning for the first
recipient is logged Postfix will not yet have allocated a queue file and
queueid for the mail, so the warning will be associated with the connection
rather than the accepted mail.  A queue file and queueid will be allocated
after Postfix accepts the MAIL FROM command if
\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
setting will cause disk IO for almost every delivery attempt, instead of
just for delivery attempts where recipients are accepted, and consequently
a drastic reduction in the performance of the mail server.

@@

This is yet another complication that only occurs during periods of
extremely high load, when process scheduling and even hard disk access
times cause strange behaviour.  In this complication bounce notification
mails are created, delivered, and deleted from the queue, \textit{before\/}
the log line from \daemon{bounce} that explains their source is logged.  To
deal with this the COMMIT action maintains a cache of recently committed
bounce notification mails, which the BOUNCE action subsequently checks if
the bounce mail is not already in the state tables.  If the queueid exists
in the cache, and its start time is less than ten seconds before the
timestamp of the bounce log line, it is assumed that the bounce
notification mail has already been processed and the BOUNCE action does not
create one.  If the queueid exists it is removed from the cache, because it
has either just been used or it is too old to use in future.  Whether the
BOUNCE action creates a mail or finds an existing mail in the state tables,
it flags the mail as having been seen by the BOUNCE action; if this flag is
present the COMMIT action will not add the mail to the cache of recent
bounce notification mails.  This is not required to correctly deal
with the complication, but is an optimisation to reduce the parser's memory
usage; on the occasions the author has observed this complication occurring
there have been a huge number of bounce notification mails generated --- if
every bounce notification mail was cached it would dramatically increase
the memory requirements of the parser.  The cache of bounce notification
mails will be pruned whenever the parser's state is saved, though if the
size of the cache ever becomes a problem it could be pruned periodically to
keep the size in check.

@@

    \item Each new Postfix release requires new rules to be written or
        existing rules modified to cope with the new or changed log lines.
        Similarly using a new \acronym{DNSBL}, a new policy server, or new
        administrator-defined rejection messages requires new rules.

@@

        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
        place to log the HELO hostname, there will not be a queueid
        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
        evaluated for the first recipient, so the log line cannot be
        associated with the correct mail.  There is guaranteed to be a
        queueid when the DATA command has been reached, and thus the
        queueid will be logged by any restrictions taking effect in
        \texttt{smtpd\_data\_restrictions}, and the log line can be
        associated with the correct mail.  There is no difficulty in
        specifying a HELO-based restriction in
        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
        correctly.

        Logging the HELO hostname in this fashion also partially prevents
        the complication described in \sectionref{Mail deleted before
        delivery is attempted} from occurring, but only when there is a
        single recipient; in that case the recipient address will be logged
        also, but when there are multiple recipients no addresses are
        logged.

@@

    \item The parser does not distinguish between log files produced by
        different sources when parsing; all results will be saved to the
        same database.  This may be viewed as an advantage, as log files
        from different sources can be combined in the same database, or it
        may be viewed as a limitation as there is no facility to
        distinguish between log files from different sources in the same
        database.  If the results of parsing log files from different
        sources must remain separate, the parser can easily be instructed
        to store the results in a different database.
